---
title: "Lab 4: Reducing Crime"
author: "Asha Anju, Anamika Sinha, Noah Randolph"
date: "8/14/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

...

## Load Data 

We first load the data and note any anomalies, such as missing values, top-coded or bottom-coded variables, or not-applicable values.

```{r load data}
library(sandwich)
library(lmtest)
library(car)
library(stargazer)

#setwd("~/Documents/W203_Statistics/Labs/Lab_4/Lab4Repository")
CrDat = read.csv("crime_v2.csv")

attach(CrDat)
```



##  Exploratory Data Analysis
We begin with a summary of the data.
```{r}
summary(CrDat)
nrow(CrDat)
ncol(CrDat)
```
##Observations: 
There are 90 observations across 25 variables. There are no missing values and the data in each variable make sense given the variable types, except for values in _probsen_ and _probconv_, which both show proportions that are greater than 1. This may be due to differences in ways the variable is reported from across the state. We will note this but would like to keep them in our analysis as they may have significant information. We also note that the maximum density is 8.8 people per square mile which indicates that we may be looking at a rural state. The crime rate proportions are low decimal values, so we will rescale it later to improve interpretation. Also, _avgsen_ has a maximum of 20 days which suggests that we have a vast majority of petty crimes. We notice a very high value of 2177.1 dollars for _wageser_ which is an extreme outlier since difference between this value and the third quartile is greater than three times the interquartile range. 

##Model proposal
Our dataset has _probsen_, _probconv_, _probarr_ , _mix_  _avgsen_, _police_  variables which represent the criminal justice system. We have variables like _wage & _tax_ which are representative of economic conditions. We also have _pctmix_, _ymale_, _density_, _urban_ variables which fall under demographic factors. There are three indicator variables representing regions(west, central) and cities(urban). 

Crime rate depends on the effectiveness of the criminal justice system and economic factors. Criminal justice systems reduce crime through deterence. Therefore increases in police, arrests, convictions, crime sentencing, and prison sentence length should reduce crime rate. Therefore, we explore these variables.

We first look at the response variable, _crime_, representing the number of each counties' crimes committed per person. The common way to report crimes is per 1000 people, so we'll do that here for readability. 

```{r fig.height = 3, fig.width = 4}
scaledcrime = crime*1000
hist(scaledcrime, breaks = 20, main = "County Crime Rates", 
     xlab = "crimes committed per 1000 people")
```

Then we take a look at other key criminal justice variables. To make police consistent with _crime_, we scale _police_ as well.

```{r}
scaledpolice = police*1000
hist(scaledpolice, breaks = 20, main = "Police Per Capita",
     xlab = "police per capita")
```

```{r}
hist(probarr, breaks = 16, 
     main = "Proportion of County Crimes Resulting in Arrests",
     xlab = "Proportion of Crimes Resulting in Arrests")
```

```{r}
hist(probconv, breaks = 10, main = "Proportion Being Convicted After Arrest",
     xlab = "Proportion")
```

```{r}
hist(probsen, breaks = 20, main = "Proportion Being Sentenced After Conviction",
     xlab = "Proportion")
```

```{r}
hist(mix, breaks = 20, main = "Ratio of Face-to-Face/All Other Crime",
     xlab = "Ratio")
```
```{r}
hist(tax, breaks = 20, main = "Tax Per Capita",
     xlab = "Tax")
```
```{r}
hist(density, breaks = 20, main = "Density",
     xlab = "people per square mile")
```

```{r}
hist(ymale, breaks = 20)
```

```{r}
hist(pctmin, breaks = 20)
```
### Observations:
Response variable
_crime_: The histogram shows a positive skew. Pesence of one outlier is noted. 

Criminal justice variables
_police_: The histogram shows a positive skew. We have one outlier which is county 115. Police per capita, _police_, should have negative influence on county crime rates, since their presence is generally thought of as a deterence.
_probarr_: The histogram shows a slight negative skew. This variable is likely to be correlated to polic since more police will lead to more arrests and vice versa.
_probsen_: The histogram shows a slight positive skew. Also, presence of two outliers is noted.  We see a probability value of greater than 1. Interestingly, this belongs to county 115, the same county which has the police outlier.
_probconv_: The histogram shows a high positive skew. We see a probability value of greater than 1 for ten counties. We note an outlier  with value 2.12 in county 185. This county also has an extreme outlier in _wageser_ of $2177.06. As mentioned before, we could not find a reasonable explaination for this. So we decided to include these in our analysis.
_mix_: The histogram shows a high positive skew. 
_avgsen_: From the initial summary, due to the low number of maximum days, we think that this variable is unlikely to deter crime.

Economic variables
Wage would be a good economic indicator but in our dataset, wage is broken up into eight different variables based on sector. In the absence of sector weights, it is difficult to create a meaningful average variable. We tried various ways to interpret the given information by taking an average of all the wage variables and also the wage range in each county to understand wage effect on crime.

_tax_: Tax revenue per capita is a good indicator of the financial health of a county. The histogram shows a positive skew. Also, presence of an outlier is noted which is county 55. We did not find anything noteworthy in this county.

Demographic variables
_density_: The histogram has a strong positive skew, reflecting a few urban areas in the state. As mentioned before, the maximum density is only eight people per square mile. 
_ymale_: This histogram also has a strong positive skew. It has an outlier in county 133. We did not find anything worth noting in this county.
_pct\_min_: The histogram has somewhat of a uniform distribution with a decrease on the right end. 


Indicator variables
_urban_: We have 8 counties in this region. 
_west_: There are 34 counties that belong to West region.
_central_: 21 counties fall in this region. 


Following the first stage of data analysis we perform a log transformation of crime and density and plot histograms

```{r}
hist(log(crime), main="Log of Crime", xlab=NULL, breaks=100)
```
```{r}
hist(log(density), main="Log of Density", xlab=NULL, breaks=100)
```

#Observations:
log(crime): The log transformation has made the distribution of log(crime) quite normal. This will help
to ensure that the errors of the model are normal (i.e. CLM 6)

log(density): The log transformation of density has helped with the normalty of the distribution  as well.

#Correlations

We examine the relationship between crime rate and potential deterrent criminal justice variables. As mentioned before, _police_ and _probarr_ are likely to be related, so we include only one out of the two. _mix_  most likely has an associative relationship to crime and will not be a deterrent. So we exclude it.

```{r}
scatterplotMatrix(~ crime + police + probconv + probsen)
scatterplotMatrix(~ log(crime) + log(police) + probconv + probsen)
```

#Observations:
Contrary to our expectation, we see a positive relationship between _police_ and _crime_. This suggests that police may be a result of high crime rate rather than a cause of low crime rate. Therefore, it will not be a key variable of interest. Both _probsen_ and _probconv_ have the expected negative relationship with _crime_, meaning that they are potentially deterrent factors. 



```{r}
scatterplotMatrix(~ crime + density + pctmin + ymale + tax)
scatterplotMatrix(~ log(crime) + log(density) + pctmin + ymale + tax)
```

#Observations:
We see a strong positive relationship between _density_ and _crime_ which is consistent with the common notion of high crime rate in urban areas. For lower _pctmin_ values, _crime_




Based on the exploratory data analysis and underlying intuition about viewing behavior an initial proposed
model specification and coefficient expectations are:
log(density):since _density_ variable ranges between 0 to 8 people per square mile..
probconv:
probsen: 

Model:



Several variables standout as good candidates for an initial model attempting to understand the determinants of crime. These include the proportion of crimes resulting in arrests, _probarr_, the average prison sentence length, _avgsen_, police per capita, _police_, people per square mile, _density_, average wages (which are spread across several variables), and the proportion of males between ages 15 and 24, _ymale_. 

The proportion of crimes resulting in arrests, _probarr_, should negatively influence crime rate, intuitively, in the same way that increased ticketing on highways should reduce speeding. The histogram and number of observations are shown below.

The histogram above shows an approximately normal distribution. With 90 observations, the central limit theorem (CLT) can be relied upon to make an inference about this variable's regression coefficient as it corresponds to the population.

The average prison sentence length, _avgsen_, like _probarr_, should also negatively influence crime rate, as longer sentences increase the consequences to committing a crime.


The prison sentence length distribution, above, is approximately normal with a positive skew, likely due to the fact that the metric is zero-bounded. As with _probarr_, 90 observations is sufficient to rely on the CLT for population inferences.







The variable describing people per square mile, _density_, is a more detailed indicator of urban versus non-urban environments than the boolean varianble, _urban_. Density should also influence crime rate, either positively or negatively, since 'haves' and 'havenots' become closer together at higher densities, yet higher density areas also result in more witnesses to deter face-to-face crime.



Density's distribution has a strong positive skew, due to the zero bound. Despite the skew, the CLT can still be relied upon for population inferences, as the number of observations, 90, is sufficient.

Average wages, _avgwage_ is a variable created below by concatenating each wage variable in the dataset together. In this way, average wages for each county can be modeled against crime.


```{r}
avgwage = (wagecon + wagetuc + wagetrd + wagefir +
        wageser + wagemfg + wagefed + wagesta + 
                wageloc)/9

length(avgwage)
hist(avgwage, breaks = 20, main = "Average Wage", xlab = "Wage")
```
Average wage also has an approximately normal distribution, with a positive skew. 
















#### Model 1

```{r}

model1 = lm(log(scaledcrime) ~ probconv + probsen + log(density) )
AIC1 = AIC(model1)
plot(model1)
summary(model1)
vif(model1)

```

#### Model 2

```{r}
model2 = lm(log(scaledcrime) ~ probconv + probsen + log(density) + scaledpolice + pctmin)
AIC2 = AIC(model2)
plot(model2)
vif(model2)
summary(model2)
```

#### Model 3

```{r}
model3 = lm(log(scaledcrime) ~ probconv + probsen +  log(density) + scaledpolice + pctmin +
                    probarr + ymale + avgsen +  mix  + central + west +
                    tax)
AIC3 = AIC(model3)
plot(model3)
vif(model3)
summary(model3)
```



```{r}
# Adjust standard errors
cov1 = vcovHC(model1)
robust_se1 = sqrt(diag(cov1))
cov2 = vcovHC(model2)
robust_se2 = sqrt(diag(cov2))
cov3 = vcovHC(model3)
robust_se3 = sqrt(diag(cov3))

# Adjust F statistic
wald_results <- waldtest(model1, vcov = cov1)

stargazer(model1, model2, model3, type = 'text', intercept.bottom = FALSE,
          se        = list(robust_se1, robust_se2, robust_se3),
          omit.stat = "f",
          add.lines = list(c("AIC", round(AIC1, 2), round(AIC2, 2),
                             round(AIC3, 2))))
```














