---
title: "Lab 4: Reducing Crime"
author: "Asha Anju, Anamika Sinha, Noah Randolph"
date: "8/14/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

...

## Load Data 

We first load the data and note any anomalies, such as missing values, top-coded or bottom-coded variables, or not-applicable values.

```{r load data}
library(sandwich)
library(lmtest)
library(car)
library(stargazer)
library(MASS)

#setwd("~/Documents/W203_Statistics/Labs/Lab_4/Lab4Repository")
CrDat = read.csv("crime_v2.csv")
attach(CrDat)
```



##  Exploratory Data Analysis
We begin with a summary of the data.
```{r}
summary(CrDat)
nrow(CrDat)
ncol(CrDat)
```
##Observations: 
There are 90 observations across 25 variables. There are no missing values and the data in each variable make sense given the variable types, except for values in _probsen_ and _probconv_, which both show proportions that are greater than 1. This may be due to differences in ways the variable is reported from across the state. We will note this but would like to keep them in our analysis as they may have significant information. We also note that the maximum density is 8.8 people per square mile which indicates that we may be looking at a rural state. The crime rate proportions are low decimal values, so we will rescale it later to improve interpretation. Also, _avgsen_ has a maximum of 20 days which suggests that we have a vast majority of petty crimes. We notice a very high value of 2177.1 dollars for _wageser_ which is an extreme outlier since difference between this value and the third quartile is greater than three times the interquartile range. 

##Model proposal
Our dataset has _probsen_, _probconv_, _probarr_ , _mix_  _avgsen_, _police_  variables which represent the criminal justice system. We have variables like _wage & _tax_ which are representative of economic conditions. We also have _pctmix_, _ymale_, _density_, _urban_ variables which fall under demographic factors. There are three indicator variables representing regions(west, central) and cities(urban). 

Crime rate depends on the effectiveness of the criminal justice system and economic factors. Criminal justice systems reduce crime through deterence. Therefore increases in police, arrests, convictions, crime sentencing, and prison sentence length should reduce crime rate. Therefore, we explore these variables.

We first look at the response variable, _crime_, representing the number of each counties' crimes committed per person. The common way to report crimes is per 1000 people, so we'll do that here for readability. 

```{r fig.height = 4, fig.width = 4}
scaledcrime = crime*1000
hist(scaledcrime, breaks = 20, main = "County Crime Rates", 
     xlab = "crimes committed per 1000 people")
```

Then we take a look at other key criminal justice variables. To make police consistent with _crime_, we scale _police_ as well.

```{r fig.height = 4, fig.width = 4}
scaledpolice = police*1000
hist(scaledpolice, breaks = 20, main = "Police Per Capita",
     xlab = "police per capita")
```

```{r fig.height = 4, fig.width = 4}
hist(probarr, breaks = 16, 
     main = "Proportion of County Crimes Resulting in Arrests",
     xlab = "Proportion of Crimes Resulting in Arrests")
```

```{r fig.height = 4, fig.width = 4}
hist(probconv, breaks = 10, main = "Proportion Being Convicted After Arrest",
     xlab = "Proportion")
```

```{r fig.height = 4, fig.width = 4}
hist(probsen, breaks = 20, main = "Proportion Being Sentenced After Conviction",
     xlab = "Proportion")
```

```{r fig.height = 4, fig.width = 4}
hist(mix, breaks = 20, main = "Ratio of Face-to-Face/All Other Crime",
     xlab = "Ratio")
```
```{r fig.height = 4, fig.width = 4}
hist(tax, breaks = 20, main = "Tax Per Capita",
     xlab = "Tax")
```
```{r fig.height = 4, fig.width = 4}
hist(density, breaks = 20, main = "Density",
     xlab = "people per square mile")
```

```{r}
hist(ymale, breaks = 20)
```

```{r}
hist(pctmin, breaks = 20)
```

### Observations:
Response variable 
_crime_: The histogram shows a positive skew. Presence of one outlier is noted. Because of the positive skew, we will do a log transformation of this variable.  
*Criminal justice variables*  
_police_: The histogram shows a positive skew. We have one outlier which is county 115. Police per capita, _police_, should have negative influence on county crime rates, since their presence is generally thought of as a deterence.  consider a log transformation of this variable. 
_probarr_: The histogram shows a slight negative skew. This variable is likely to be correlated to polic since more police will lead to more arrests and vice versa. 
_probsen_: The histogram shows a slight positive skew. Also, presence of two outliers is noted.  We see a probability value of greater than 1. Interestingly, this belongs to county 115, the same county which has the police outlier. 
_probconv_: The histogram shows a high positive skew. We see a probability value of greater than 1 for ten counties. We note an outlier  with value 2.12 in county 185. This county also has an extreme outlier in _wageser_ of $2177.06. As mentioned before, we could not find a reasonable explaination for this. So we decided to include these in our analysis. 
_mix_: The histogram shows a high positive skew. 
_avgsen_: From the initial summary, due to the low number of maximum days, we think that this variable is unlikely to deter crime. 

Economic variables 
Wage would be a good economic indicator but in our dataset, wage is broken up into eight different variables based on sector. In the absence of sector weights, it is difficult to create a meaningful average variable. We tried various ways to interpret the given information by taking an average of all the wage variables and also the wage range in each county to understand wage effect on crime.

_tax_: Tax revenue per capita is a good indicator of the financial health of a county. The histogram shows a positive skew. Also, presence of an outlier is noted which is county 55. We did not find anything noteworthy in this county.

Demographic variables
_density_: The histogram has a strong positive skew, reflecting a few urban areas in the state. As mentioned before, the maximum density is only eight people per square mile. 
_ymale_: This histogram also has a strong positive skew. It has an outlier in county 133. We did not find anything worth noting in this county.
_pct\_min_: The histogram has somewhat of a uniform distribution with a decrease on the right end. 


Indicator variables
_urban_: We have 8 counties in this region. 
_west_: There are 34 counties that belong to West region.
_central_: 21 counties fall in this region. 


Following the first stage of data analysis we perform a log transformation of crime and density and plot histograms.

```{r}
hist(log(scaledcrime), main="Log of Crime", xlab=NULL, breaks=20)
```
```{r}
hist(log(density), main="Log of Density", xlab=NULL, breaks=20)
```

#Observations:
_log(scaledcrime)_: The log transformation has made the distribution quite normal. This will help ensure that the errors of the model are normal (i.e. CLM assumption 6).
_log(density)_: The log transformation of density has helped with the normalty of the distribution  as well.

#Correlations
We examine the relationship between crime rate and potentially deterrent criminal justice variables. As mentioned before, _police_ and _probarr_ are likely to be highly correlated, since police are the ones making arrests, so we include only one out of the two. _mix_ most likely has an associative relationship to crime and will not be a deterrent, so we exclude it.

```{r}
scatterplotMatrix(~ log(scaledcrime) + log(scaledpolice) + probconv + probsen)
```

#Observations:
Contrary to our expectation, we see a positive relationship between _log(scaledpolice)_ and _log(scaledcrime)_. This suggests that police may be the result of high crime rate rather than a cause of low crime rate. Therefore, it will not be a key variable of interest. Both _probsen_ and _probconv_ have the expected negative relationship with crime rate, meaning that they are potentially deterrent factors. 

```{r}
scatterplotMatrix(~ log(scaledcrime) + log(density) + pctmin + log(ymale) + log(tax))
```

##Observations:
We see a strong positive relationship between _log(density)_ and _log(scaledcrime)_, which is consistent with the common notion of high crime rate in urban areas. The association between _pctmin_ and _log(scaledcrime)_ is less clear, with an initial rise, then a reduction as _pctmin_ increases on the x-axis. Even after the log transformation, _log(ymale)_ still has a high positive skew and therefore too much clustering to make sense of the relationship with crime rate. The same is true for _log(tax)_. Furthermore, we expected to see a negative relationship with tax revenue per capita as a proxy for income levels. Since it's positive, we rule out tax as a key variable of interest as it relates to reducing crime rates.

#### Model 1
Based on the exploratory data analysis and underlying intuition about crime relationships, our initial proposed model specification and coefficient expectations are:
_log(density)_: since this variable shows the strongest linear relationship with _log(scaledcrime)_ and is consistent across its range, we include this as an associative variable to avoid significant omitted variable bias in our key variables of interest.
_probconv_ and _probsen_: our intuition that an increase in conviction and sentencing rates helps reduce crime rate is supported by the scatterplot in our EDA.

```{r}
model1 = lm(log(scaledcrime) ~ probconv + log(density))
```

#### Testing the validity of the 6 assumptions of the CLM
**CLM 1 - A linear model**
The model is specified such that the dependent variable is a linear function of the explanatory variables. 
Is the assumption valid? Yes.
Response: No response required.

**CLM 2 - Random Sampling**
Given that we have a dataset of 90 counties from an unknown location, and we do not know how the data was collected and if the data collection was consistent among the counties, this assumption is uncertain. Additionally the structure of one observation representing each county further confounds the assumption of randomly observed variables. A true random sample would take observations at random from the unaggregated level. 
Is the assumption valid? Not clear.
Response: We proceed with the analysis.

**CLM 3 - Multicollinearity**
As a quick test of the multicollinearity condition, we check the correlation of the three explanatory variables and their Variance Inflation Factors (VIF):

```{r}
CrDat$logdensity = log(density)
X = data.matrix(subset(CrDat, select=c(logdensity, probconv, probsen)))
cor(X)
vif(model1)
```

The three explanatory variables are not perfectly correlated and the VIFs are low (i.e. less than 4), so there is no perfect multicollinearity of the independent variables.
Is the assumption valid? Yes 
Response: No response required.

***CLM 4 – Zero-Conditional Mean***
To analyze whether there is a zero-conditional mean of the residuals across the explanatory variables in model 1, we plot the residuals against the fitted values with the predicted conditional mean spline line across fitted values.
```{r}
plot(model1, which=1)
```

The plot indicates little evidence that the zero-conditional mean assumption doesn’t hold. The red spline line on the residuals versus fitted values plot is fairly flat, with slight trends away from zero toward the edges. Given that there are few data points at the edges, the slight trends are not a concern.
Is the assumption valid? Yes
Response: No response required

**CLM 5 - Homoscedasticity**
To determine whether the variance of the error is fixed for all model 1 explanatory variables, we view the scale-location plot. 

```{r}
plot(model1, which = 3)
bptest(model1)
```

The scale-location plot shows considerable evidence of heteroskedasticity, as seen by the considerable red spline line curvature. Also the Breusch-Pagan shows a rejection of the null hyposthesis of homoskedasticity, with a p-value less than \alpha = 0.01.

Is the assumption valid? No.
Response: use robust standard errors.

**CLM 6 – Normality of residuals**
To determine whether there is normality of the residuals we use a Q-Q plot of the residuals and simply visually observe whether there is normality.

```{r}
plot(model1, which = 2)
```

The Q-Q plot shows deviation from normality on both the negative end and, considerably more, at the positive end. 

Is the assumption valid? No.
Response: Despite the distribution of model 1 residuals showing non-normality, since we have a large sample size of 90, we can rely on the central limit theorem to satisfy this assumption.

#### Model 2

```{r}
model2 = lm(log(scaledcrime) ~ probconv + probsen + log(density) + pctmin)
AIC2 = AIC(model2)
plot(model2)
vif(model2)
summary(model2)
```

#### Model 3

```{r}
model3 = lm(log(scaledcrime) ~ probconv + probsen +  log(density) + pctmin + probarr + ymale + avgsen +  mix  + central + west + tax)
AIC3 = AIC(model3)
plot(model3)
vif(model3)
summary(model3)
```



```{r}
# Compute Akaike information criterion (AIC)
AIC1 = AIC(model1)
AIC2 = AIC(model2)
AIC3 = AIC(model3)

# Adjust standard errors
cov1 = vcovHC(model1)
robust_se1 = sqrt(diag(cov1))
cov2 = vcovHC(model2)
robust_se2 = sqrt(diag(cov2))
cov3 = vcovHC(model3)
robust_se3 = sqrt(diag(cov3))

# Adjust F statistic
wald_results <- waldtest(model1, vcov = cov1)

stargazer(model1, model2, model3, type = 'text', intercept.bottom = FALSE,
          se        = list(robust_se1, robust_se2, robust_se3),
          omit.stat = "f",
          add.lines = list(c("AIC", round(AIC1, 2), round(AIC2, 2),
                             round(AIC3, 2))))
```


******************************
Several variables standout as good candidates for an initial model attempting to understand the determinants of crime. These include the proportion of crimes resulting in arrests, _probarr_, the average prison sentence length, _avgsen_, police per capita, _police_, people per square mile, _density_, average wages (which are spread across several variables), and the proportion of males between ages 15 and 24, _ymale_. 

The proportion of crimes resulting in arrests, _probarr_, should negatively influence crime rate, intuitively, in the same way that increased ticketing on highways should reduce speeding. The histogram and number of observations are shown below.

The histogram above shows an approximately normal distribution. With 90 observations, the central limit theorem (CLT) can be relied upon to make an inference about this variable's regression coefficient as it corresponds to the population.

The average prison sentence length, _avgsen_, like _probarr_, should also negatively influence crime rate, as longer sentences increase the consequences to committing a crime.


The prison sentence length distribution, above, is approximately normal with a positive skew, likely due to the fact that the metric is zero-bounded. As with _probarr_, 90 observations is sufficient to rely on the CLT for population inferences.







The variable describing people per square mile, _density_, is a more detailed indicator of urban versus non-urban environments than the boolean varianble, _urban_. Density should also influence crime rate, either positively or negatively, since 'haves' and 'havenots' become closer together at higher densities, yet higher density areas also result in more witnesses to deter face-to-face crime.



Density's distribution has a strong positive skew, due to the zero bound. Despite the skew, the CLT can still be relied upon for population inferences, as the number of observations, 90, is sufficient.

Average wages, _avgwage_ is a variable created below by concatenating each wage variable in the dataset together. In this way, average wages for each county can be modeled against crime.


```{r}
avgwage = (wagecon + wagetuc + wagetrd + wagefir +
        wageser + wagemfg + wagefed + wagesta + 
                wageloc)/9

length(avgwage)
hist(avgwage, breaks = 20, main = "Average Wage", xlab = "Wage")
```
Average wage also has an approximately normal distribution, with a positive skew. 
********************************************************












